---
title: "Baseball Statistics"
author: "Stryder, Padraic, Chesteran, Shyan"
date: "5/8/2021"
output:
  word_document: default
  html_document:
    df_print: paged
---

 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##

## Introduction

The title of our database is called Baseball Databank. In our project, we will be analyzing our Baseball database and writing brief documentation about it. The data set contains Excel sheets consisting of all the baseball statistical data from the years 1871-2015, we will only be using the data from 1995 to 2015; and will be including 7 data frames. The first dataframe is batting which contains statistics related to batting. The second data frame battingPost, contains batting statistics in the postseason. The third Fielding, contains variables related to fielding statistics. The fourth Pitching, contains pitching statistics for a given player on a given year. The fifth pitchingPost, contains statistics for pitching in the postseason. The sixth master, contains data about each player’s name, date of birth, and biographical information in the dataset. And lastly Salaries, contains a player’s salary in a given year. Every data frame is linked together with PlayerID. 

For our dataset, we are going to answer a series of 6 questions that we believe our dataset can answer. First, we want to know if there is a correlation between salaries and the performance of players? To determine the correlation we are going to compare player’s WAR to their salaries.  Second, we want to know how the pressure of the postseason impacts a player’s performance. We will solve this by comparing a player’s WAR in the regular season to a player’s WAR in the postseason. Third, we want to determine the shortfalls of using traditional baseball statistics to rate a player’s performance. In order to figure out this question, we will compare an advanced statistic such as wOBA and compare it with a traditional statistical counterpart such as Batting Average. We will also compare players who perform well according to traditional baseball statistics and see if the advanced statistics tell the same story. Fourth, we want to determine if there is a correlation between how much a player is paid and the individual accolades he earns by comparing the salaries of award winners. Fifth, we want to determine the correlation between the player’s age and his overall performance by examining the relationship between age and WAR. Finally, we want to determine whether birth location has an influence on how a player performs and to solve this we will compare  birth location to WAR.

## Intial Goals

  1.    Initial Goals
  2.    Can we predict the number of runs a player will score over a season?
  3.    Can we predict a position player's performance over a season? 
  4.    Can we predict a pitcher's performance over a season? 
  5.    What are the flaws of traditional baseball statistics?
  6.    Is salary a good predictor of Awards won?
  7.    Can we predict the number of earned runs (ER) a pitcher will allow over a season?
  8.    Can we predict a position player’s batting postseason performance by using their regular season data?

## Library used

```{r, echo=FALSE}
library(ggbeeswarm)
library(ggforce)
library(readr)
library(tidyverse)
library(dplyr)
library(caret)
library(rpart)
library(car)
library(e1071)
library(kernlab)
library(randomForest)
library(sjmisc)
library(lme4)
library(ModelMetrics)
library(knitr)
```
## Importing the data frames
```{r, echo=FALSE}
AwardsPlayers <- read_csv("AwardsPlayers.csv")
Batting <- read_csv("Batting.csv")
BattingPost <- read_csv("BattingPost.csv")
Fielding <- read_csv("Fielding.csv")
Master <- read_csv("Master.csv")
Pitching <- read_csv("Pitching.csv")
PitchingPost <- read_csv("PitchingPost.csv")
Salaries <- read_csv("Salaries.csv")
```
## Filtering the data set
```{r, echo=FALSE}
##Filtering dataframes to only include data between 1995-2015 
AwardsPlayers <- AwardsPlayers %>% filter(yearID >= 1995)
Batting <- Batting %>% filter(yearID >= 1995)
BattingPost <- BattingPost %>% filter(yearID >= 1995)
Fielding <- Fielding %>% filter(yearID >= 1995)
Pitching <- Pitching %>% filter(yearID >= 1995)
PitchingPost <- PitchingPost %>% filter(yearID >= 1995)
Master <- Master %>% filter(finalGame >= as.Date("1995-04-25"))
Salaries <- Salaries %>% filter(yearID >= 1995)
```
## Transforming the Batting Statistics
```{r, echo=FALSE}
## Transforming Batting Statistics
## removing undesirable variable
Batting2 <- Batting %>% subset(select = -c(teamID, lgID, SF, GIDP, stint))
## na is equivalent to 0
Batting2[is.na(Batting2)] = 0
## Combining all player batting stats into seasons
Batting3 <- aggregate(. ~ playerID + yearID, data = Batting2, sum)
## Creating on base average (OBA) higher indicates better players
Batting3 <- Batting3 %>% mutate(OBA = ((H + BB + HBP)/(AB+ BB + HBP)))
## Creating extra base hits (XBH)
Batting3 <- Batting3 %>% mutate(XBH = (`2B` + `3B` + HR))
## Creating Slugging (SLG) higher indicates better players
Batting3 <- Batting3 %>% mutate(SLG = (((H-XBH)+(2*`2B`)+(3*`3B`)+(4*HR))/(AB)))
## Creating On Base Plus Slugging (OPS) higher indicates better players
Batting3 <- Batting3 %>% mutate(OPS = (OBA + SLG))
## Creating on base average Stolen Base Percentage (SBP) higher indicates better players
Batting3 <- Batting3 %>% mutate(SBP = ((SB)/(SB + CS)))
## Creating on singles stat
Batting3 <- Batting3 %>% mutate(`1B` = (H-XBH))
## Creating on singles stat
Batting3 <- Batting3 %>% mutate(BO = (AB - H))  
## Creating BA
Batting3 <- Batting3 %>% mutate(BA = (AB/H))
## Creating wOBA
Batting3 <- Batting3 %>% mutate(wOBA = (.69*(BB) +.72*(HBP) +.89*(H - XBH) + 1.27*(`2B`) + 1.62*(`3B`) + HR*2.10) / (AB + BB - IBB + SH +HBP))
```
## Transforming the Fielding (Position Players) Statistics
```{r, echo=FALSE}
##Transforming fielding statistics
## removing undesirable variables 
SFielding2 <- Fielding %>% subset(select = -c(teamID, lgID,InnOuts, WP, SB, CS, ZR, GS, stint))
## removing pitchers from fielding
SFielding2 <- SFielding2 %>% filter(POS != "P")
## removing position
SFielding2 <- SFielding2 %>% subset(select = -c(POS))
## na is equivalent to 0
SFielding2[is.na(SFielding2)] = 0
## Combining all player fielding stats into seasons
SFielding3 <- aggregate(. ~  playerID + yearID, data = SFielding2, sum)
## Combines the data frames batting and fielding together contain only position player observations 
## creating Fielding Percentage(FP)
SFielding3 <- SFielding3 %>% mutate(FP = ((PO + A)/(PO + A + E)))
```
## Tranforming the Pitching Statistics
```{r, echo=FALSE}
## removing non numeric variables in pitching
pitching2 <- Pitching %>% subset(select = -c(teamID, lgID, IBB, SH, SF, GIDP, ERA))
## combining all the stats from the different stints of the pitchers
pitching2 <- aggregate(. ~ playerID + yearID, data = pitching2, sum)
##Creating a combined ERA
pitching2 <- pitching2 %>% mutate(ERA = ((ER *9)/(IPouts/3)))
## removing the stint column
pitching2 <- pitching2 %>% subset(select = -c(stint))
## removing non numeric variables in fielding
fielding2 <- Fielding %>% subset(select = -c(teamID, lgID, GS, InnOuts,PB, WP, SB, CS, ZR))
## filtering out all non pitchers
fielding2 <- fielding2 %>% filter(POS == "P")
## removing the position column
fielding2 <- fielding2 %>% subset(select = -c(POS))
## combining all the stats from the different stints of the pitchers
fielding2 <- aggregate(. ~ playerID + yearID, data = fielding2, sum)
## removing the stint column
fielding2 <- fielding2 %>% subset(select = -c(stint))
## complete data frame containing all statistics recorded for pitchers
pitcher_stats <- left_join(fielding2, pitching2, by = c("playerID", "yearID"))
## removing the playerID column
pitcher_stats2 <- pitcher_stats %>% subset(select = -c(playerID))
## combining everything by year and averaging all the states league wide
pitcher_stats2 <- aggregate(. ~ yearID, data = pitcher_stats2, mean)
## removing the all non important columns
pitcher_stats3 <- pitcher_stats2 %>% subset(select = c(yearID, ERA, HR, BB, HBP, SO, IPouts))
## converting IPouts to IP
pitcher_stats3$IPouts <- pitcher_stats3$IPouts/3
##renaming all the columns
pitcher_stats3 <- pitcher_stats3 %>% rename(lgERA = ERA, lgHR = HR, lgBB = BB, lgHBP = HBP, lgSO = SO, lgIP = IPouts)
##calculating the FIP constant
pitcher_stats3 <- pitcher_stats3 %>% mutate(FIPconstant =  lgERA - (((13*lgHR)+(3*(lgBB+lgHBP))-(2*lgSO))/lgIP))
## create new dataframe with all the league avg states and the FIP constant per year
pitcher_stats_complete <- left_join(pitcher_stats, pitcher_stats3, by = c("yearID"))
## removing G.y and renaming G.x
pitcher_stats_complete <- pitcher_stats_complete %>% subset(select = -c(G.y))
pitcher_stats_complete <- pitcher_stats_complete %>% rename(G = G.x)
## converting and renaming IPouts to IP
pitcher_stats_complete$IPouts <- pitcher_stats_complete$IPouts/3
pitcher_stats_complete <- pitcher_stats_complete %>% rename(IP = IPouts)
## calculating the FIP for every player  
pitcher_stats_complete <- pitcher_stats_complete %>% mutate(FIP = ((13*HR)+(3*(BB+HBP))-(2*SO))/IP + FIPconstant)
##filtering out pitchers with low playing time
pitcher_stats_complete <- pitcher_stats_complete %>% filter(IP >= 20)
##filtering out infinites 
pitcher_stats_complete <- pitcher_stats_complete %>% filter(is.finite(FIP))
```
## Creating WAR
```{r, echo=FALSE}
## Contains both fielding and batting statistics for each player by season
Player_WAR <- full_join(Batting3, SFielding3, by = c("playerID", "yearID"))  
## Removing undesirable variables
##Player_WAR <- Player_WAR %>% subset(select = c(playerID, yearID, OPS, SBP, FP, G.x))
## Calculating the means
OPSmean <- mean(Player_WAR$OPS, na.rm ="TRUE")
SBPmean <- mean(Player_WAR$SBP, na.rm = "TRUE")
FPmean <- mean(Player_WAR$FP, na.rm = "TRUE")
## Creating a statistic that determines value of player based on fielding, batting, and base running
Player_WAR <- Player_WAR %>% mutate(WAR = ((OPS/OPSmean)+(SBP/SBPmean)+(FP/FPmean)))
```
## Batting Post
```{r, echo = FALSE}
##Comparing Regular VS Postseason OPS
## removing undesirable variable
BattingPost2 <- BattingPost %>% subset(select = -c(teamID, lgID, SF, GIDP, round))
## na is equivalent to 0
BattingPost2[is.na(BattingPost2)] = 0
## Combining all player batting stats into seasons
BattingPost3 <- aggregate(. ~ playerID + yearID, data = BattingPost2, sum)
## Creating on base average (OBA)
BattingPost3 <- BattingPost3 %>% mutate(OBA = ((H + BB + HBP)/(AB+ BB + HBP)))
## Creating extra base hits (XBH)
BattingPost3 <- BattingPost3 %>% mutate(XBH = (`2B` + `3B` + HR))
## Creating Slugging (SLG)
BattingPost3 <- BattingPost3 %>% mutate(SLG = (((H-XBH)+(2*`2B`)+(3*`3B`)+(4*HR))/(AB)))
## Creating On Base Plus Slugging (OPS)
BattingPost3 <- BattingPost3 %>% mutate(OPS = (OBA + SLG))
## Creating on singles stat
BattingPost3 <- BattingPost3 %>% mutate(`1B` = (H-XBH))
## Creating on singles stat
BattingPost3 <- BattingPost3 %>% mutate(BO = (AB - H))
```
## Functions
```{r, echo =FALSE}
rmse <- function(error)
{
  sqrt(mean(error^2))
}
RSQUARE <- function(y_actual,y_predict){
  cor(y_actual,y_predict)^2
}
```
##  Question 1 Predicting Runs (R)
```{r}
##WAR vs. Salary
WAR_Salary <- left_join(Player_WAR, Salaries,by = c("playerID", "yearID"))
WAR_Salary <- WAR_Salary %>% rename(G = G.x)
WAR_Salary2 <- na.omit(WAR_Salary)
WAR_Salary2 <- WAR_Salary2 %>% subset(select = -c(playerID, yearID))
WAR_Salary2 <- WAR_Salary2 %>% filter(AB > 100)
WAR_Salary2 <- WAR_Salary2 %>% filter(G > 20)
```
## Visualization
```{r, echo =FALSE}
ggplot(data = WAR_Salary2, mapping = aes(x = AB, y = R, color = wOBA)) + geom_point() + geom_smooth(color = "red")
ggplot(data = WAR_Salary2, mapping = aes(x = H, y = R, color = wOBA)) + geom_point() + geom_smooth(color = "red")
```
## partitioning the data
```{r, echo=FALSE}
R.ts <- WAR_Salary2$R %>% createDataPartition(p = .8, list = FALSE)
R.train <- WAR_Salary2[R.ts,]
R.test <- WAR_Salary2[-R.ts,]
```
## training data

## MLR
```{r}
modelMLRR <- lm(R ~ `1B` + `2B` + `3B` + HR + BB + HBP + SB + CS + salary,data = R.train)
vif(modelMLRR)
summary(modelMLRR)
predictedR <- predict(modelMLRR, R.train)
RMSE(predictedR, R.train$R)
R2(predictedR, R.train$R)
```
## SVR
```{r}
modelSVRR <- svm(R ~ `1B` + `2B` + `3B` + HR + BB + HBP + SB + CS + salary , data = R.train)
predictedR2 <- predict(modelSVRR, R.train)
RMSE(predictedR2, R.train$R)
R2(predictedR2, R.train$R)
```
## DT
```{r}
fitR <- rpart( R ~ `1B` + `2B` + `3B` + HR + BB + HBP + SB + CS + salary, data = R.train, method = "anova")
plot(fitR, uniform = TRUE,
     main = "Runs Decision Tree using Regression")
dev.off()
print(fitR)
summary(fitR)
predictedR3 <- predict(fitR, R.train, method = "anova")
RMSE(predictedR3, R.train$R)
R2(predictedR3, R.train$R)
```
## testing data

## MLR
```{r}
predictedR4 <- predict(modelMLRR, R.test)
RMSE(R.test$R,predictedR4)
R2(R.test$R,predictedR4)
```
## SVR
```{r}
predictedR5 <- predict(modelSVRR, R.test)
RMSE(predictedR5, R.test$R)
R2(predictedR5, R.test$R)
```
## DT
```{r}
predictedR6 <- predict(fitR, R.test, method = "anova")
RMSE(predictedR6, R.test$R)
R2(predictedR6, R.test$R)
```









##Question 2 Position Position Player performance

##WAR vs Salary
##visualization
```{r}
ggplot(data = WAR_Salary, mapping = aes(x = salary, y = WAR, color = AB)) + geom_point() + geom_smooth()
```

##modeling
##MLR
```{r, echo=FALSE}
WAR_Salary2 <- na.omit(WAR_Salary)
WAR_Salary2 <- WAR_Salary2 %>% subset(select = -c(playerID, yearID))
WAR_Salary2 <- WAR_Salary2 %>% filter(AB > 100)
WAR_Salary2 <- WAR_Salary2 %>% filter(G > 20)

set.seed(123)
training.samples <- WAR_Salary2$WAR %>% createDataPartition(p = 0.8, list = FALSE)
train.data <- WAR_Salary2[training.samples, ]
test.data <- WAR_Salary2[-training.samples, ]
```

```{r}
model1b <- lm(WAR ~ SB + CS + SO + IBB + HBP + SH + PB + FP + H, data = train.data)
summary(model1b)

##train Data
predictedR <- predict(model1b, train.data)
RMSE(predictedR, train.data$WAR)
R2(predictedR, train.data$WAR)

##test Data
predictedR <- predict(model1b, test.data)
RMSE(predictedR, test.data$WAR)
R2(predictedR, test.data$WAR)
```

##SVR
```{r}
model_1_SVR <- svm(WAR ~ SB + CS + SO + IBB + HBP + SH + PB + FP + H, data = train.data)

##Train Data
predicted.classes <- predict(model_1_SVR, train.data)
RMSE(predicted.classes, train.data$WAR)
R2(predicted.classes, train.data$WAR)

##Test Data
predicted.classes <- predict(model_1_SVR, test.data)

RMSE(predicted.classes, test.data$WAR)
R2(predicted.classes, test.data$WAR)
```

##DT
```{r}
DT_1_WAR <- rpart(WAR ~ SB + CS + SO + IBB + HBP + SH + PB + FP + H, data = train.data, method = "anova")

plot(DT_1_WAR, uniform = TRUE,
     main = "Runs Decision Tree using Regression")
dev.off()
print(DT_1_WAR)
summary(DT_1_WAR)

##train
predictedR3 <- predict(DT_1_WAR, train.data, method = "anova")
RMSE(predictedR3, train.data$WAR)
R2(predictedR3, train.data$WAR)

##test
predictedR3 <- predict(DT_1_WAR, test.data, method = "anova")
RMSE(predictedR3, test.data$WAR)
R2(predictedR3, test.data$WAR)
```

## WAR vs AGE
```{r}
## removing all unnecessary columns in master
master1 <- Master %>%  subset(select = c("playerID", "birthYear", "birthCountry", "birthState", "birthCity", "nameFirst", "nameLast", "nameGiven"))

## adding the contents of master to the pitching stats and calculating the age for every player for every year
pitcher_stats_complete_AGE <- left_join(pitcher_stats_complete, master1, by = c("playerID"))
pitcher_stats_complete_AGE <- pitcher_stats_complete_AGE %>% mutate(age = yearID-birthYear)

## adding the contents of master to the player war and calculating the age for every player for every year
Player_WAR_AGE <- left_join(Player_WAR, master1, by = c("playerID"))
Player_WAR_AGE <- Player_WAR_AGE %>% mutate(age = yearID-birthYear)
```

##visualization
```{r}
ggplot(data = Player_WAR_AGE) + geom_smooth(mapping = aes(x = age, y = WAR))
```

##MLR
```{r}
Player_WAR_AGE <- na.omit(Player_WAR_AGE)

set.seed(123)
training.samples <- Player_WAR_AGE$WAR %>% createDataPartition(p = 0.8, list = FALSE)
train.data <- Player_WAR_AGE[training.samples, ]
test.data <- Player_WAR_AGE[-training.samples, ]
```

```{r}
MLR_5_WAR <- lm(WAR ~ H + SB + CS + SH + SLG + PB + age , data = train.data)
summary(MLR_5_WAR)

##train Data
predictedR <- predict(model1b, train.data)
RMSE(predictedR, train.data$WAR)
R2(predictedR, train.data$WAR)

##test Data
predictedR <- predict(model1b, test.data)

RMSE(predictedR, test.data$WAR)
R2(predictedR, test.data$WAR)
```

##SVR
```{r}
SVR_5_WAR <- svm(WAR ~ H + SB + CS + SH + SLG + PB + age , data = train.data)

##Train Data
predicted.classes <- predict(SVR_5_WAR, train.data)
RMSE(predicted.classes, train.data$WAR)
R2(predicted.classes, train.data$WAR)

##Test Data
predicted.classes <- predict(SVR_5_WAR, test.data)

RMSE(predicted.classes, test.data$WAR)
R2(predicted.classes, test.data$WAR)
```

##DT
```{r}
DT_5_WAR <- rpart(WAR ~ H + SB + CS + SH + SLG + PB + age, data = train.data, method = "anova")

plot(DT_5_WAR, uniform = TRUE,
     main = "Runs Decision Tree using Regression")
dev.off()
print(DT_5_WAR)
summary(DT_5_WAR)

##train
predictedR3 <- predict(DT_5_WAR, train.data, method = "anova")
RMSE(predictedR3, train.data$WAR)
R2(predictedR3, train.data$WAR)

##test
predictedR3 <- predict(DT_1_WAR, test.data, method = "anova")
RMSE(predictedR3, test.data$WAR)
R2(predictedR3, test.data$WAR)
```

##WAR vs birthState
```{r}
stateavgFIP <- pitcher_stats_complete_AGE %>% subset(select = c(birthCountry, birthState, FIP))
stateavgWAR <- Player_WAR_AGE %>% subset(select = c(birthCountry, birthState, WAR))

stateavgFIP <- stateavgFIP %>% filter(birthCountry == "USA")
stateavgWAR <- stateavgWAR %>% filter(birthCountry == "USA")

stateavgFIP <- stateavgFIP %>% subset(select = -c(birthCountry))
stateavgWAR <- stateavgWAR %>% subset(select = -c(birthCountry))

stateavgFIP <- stateavgFIP %>% filter(is.finite(FIP))
stateavgWAR <- stateavgWAR %>% filter(is.finite(WAR))

stateavgFIP <- aggregate(. ~ birthState, data = stateavgFIP, mean)
stateavgWAR <- aggregate(. ~ birthState, data = stateavgWAR, mean)

stateFIP <- pitcher_stats_complete_AGE
stateWAR <- Player_WAR_AGE

stateFIP <- stateFIP %>% filter(birthCountry == "USA")
stateWAR <- stateWAR %>% filter(birthCountry == "USA")

stateFIP$birthState <- as.factor(stateFIP$birthState)
stateWAR$birthState <- as.factor(stateWAR$birthState)

stateFIP <- na.omit(stateFIP)
stateWAR <- na.omit(stateWAR)
```

##visualization
```{r}
ggplot(data=stateavgWAR, aes(x=birthState, y=WAR)) + geom_bar(stat="identity", color="blue", fill="white")
```

##modeling
##MLR
```{r}
set.seed(123)
training.samples <- stateWAR$WAR %>% createDataPartition(p = 0.8, list = FALSE)
train.data <- stateWAR[training.samples, ]
test.data <- stateWAR[-training.samples, ]
```

```{r}
MVR_6_WAR <- lm(WAR ~ AB + R + SB + CS + SH + OBA + SLG + birthState, data = train.data)
summary(MVR_6_WAR)

##train Data
predictedR <- predict(MVR_6_WAR, train.data)

RMSE(predictedR, train.data$WAR)
R2(predictedR, train.data$WAR)

##test Data
predictedR <- predict(MVR_6_WAR, test.data)

RMSE(predictedR, test.data$WAR)
R2(predictedR, test.data$WAR)
```

##SVR
```{r}
SVR_6_WAR <- svm(WAR ~ AB + R + SB + CS + SH + OBA + SLG + birthState, data = train.data)

##Train Data
predicted.classes <- predict(SVR_6_WAR, train.data)
RMSE(predicted.classes, train.data$WAR)
R2(predicted.classes, train.data$WAR)

##Test Data
predicted.classes <- predict(SVR_6_WAR, test.data)

RMSE(predicted.classes, test.data$WAR)
R2(predicted.classes, test.data$WAR)
```

##DT
```{r}
DT_6_WAR <- rpart(WAR ~ AB + R + SB + CS + SH + OBA + SLG + birthState, data = train.data, method = "anova")

plot(DT_6_WAR, uniform = TRUE,
     main = "Runs Decision Tree using Regression")
dev.off()
print(DT_6_WAR)
summary(DT_6_WAR)

##train
predictedR3 <- predict(DT_6_WAR, train.data, method = "anova")
RMSE(predictedR3, train.data$WAR)
R2(predictedR3, train.data$WAR)

##test
predictedR3 <- predict(DT_6_WAR, test.data, method = "anova")
RMSE(predictedR3, test.data$WAR)
R2(predictedR3, test.data$WAR)
```

##Question 3
##FIP vs Salary

##visualization
```{r}
FIP_Salary <- left_join(pitcher_stats_complete, Salaries,by = c("playerID", "yearID"))
##Adding age
##filtering out small sample sizes (games)
FIP_Salary <- FIP_Salary %>% filter(IP > 30)
##filtering out outliers(WAR)
FIP_Salary <- FIP_Salary %>% filter(FIP < 9)
##Visualization
ggplot(data = FIP_Salary, mapping = aes(x = salary, y = FIP, color = IP)) + geom_point() +geom_smooth()

```
##modeling
```{r}
FIP_Salary2 <- FIP_Salary %>% subset(select = -c(playerID, yearID, teamID, lgID))

set.seed(123)
training.samples <- FIP_Salary2$FIP %>% createDataPartition(p = 0.8, list = FALSE)
train.data <- FIP_Salary2[training.samples, ]
test.data <- FIP_Salary2[-training.samples, ]
```

##MLR
```{r}
model1c <- lm(FIP ~ PO + E + DP + W + L + CG + HR + BB + BAOpp + WP + HBP + ERA, data = train.data) 
summary(model1c)

##train Data
predictedR <- predict(model1c, train.data)
RMSE(predictedR, train.data$FIP)
R2(predictedR, train.data$FIP)

##test Data
predictedR <- predict(model1c, test.data)
RMSE(predictedR, test.data$FIP)
R2(predictedR, test.data$FIP)

```
##DT
```{r}
DT_1_FIP <- rpart(FIP ~ PO + E + DP + W + L + CG + HR + BB + BAOpp + WP + HBP + ERA, data = train.data, method = "anova")

plot(DT_1_FIP, uniform = TRUE,
     main = "Runs Decision Tree using Regression")
dev.off()
print(DT_1_FIP)
summary(DT_1_FIP)

##Train Data
predictedR3 <- predict(DT_1_FIP, train.data, method = "anova")
RMSE(predictedR3, train.data$FIP)
R2(predictedR3, train.data$FIP)

##Test Data
predictedR <- predict(DT_1_FIP, data = train.data, method = "anova")
RMSE(predictedR, train.data$FIP)
R2(predictedR, train.data$FIP)
```


##FIP vs Age
##visualization
```{r}
ggplot(data = pitcher_stats_complete_AGE) + geom_smooth(mapping = aes(x = age, y = FIP))
```

##Modeling
##MLR
```{r}
set.seed(123)
training.samples <- pitcher_stats_complete_AGE$FIP %>% createDataPartition(p = 0.8, list = FALSE)
train.data1 <- pitcher_stats_complete_AGE[training.samples, ]
test.data1 <- pitcher_stats_complete_AGE[-training.samples, ]
```

```{r}
MVR_5_FIP <- lm(FIP ~ PO + A + E + DP + W + L + CG + HR + BB + BAOpp + WP + HBP + BK + ERA, data = train.data)

##Train Data
predicted.classes <- predict(MVR_5_FIP, train.data)

RMSE(predicted.classes, train.data$FIP)
R2(predicted.classes, train.data$FIP)

##Test Data
predicted.classes <- predict(MVR_5_FIP, test.data)

RMSE(predicted.classes, test.data$FIP)
R2(predicted.classes, test.data$FIP)

```
##DT
```{r}
DT_5_FIP <- rpart(FIP ~ PO + A + E + DP + W + L + CG + HR + BB + BAOpp + WP + HBP + BK + ERA, data = train.data, method = "anova")

plot(DT_5_FIP, uniform = TRUE,
     main = "Runs Decision Tree using Regression")
dev.off()
print(DT_5_FIP)
summary(DT_5_FIP)

##train
predictedR3 <- predict(DT_5_FIP, train.data, method = "anova")
RMSE(predictedR3, train.data$FIP)
R2(predictedR3, train.data$FIP)

##test
predictedR3 <- predict(DT_5_FIP, test.data, method = "anova")
RMSE(predictedR3, test.data$FIP)
R2(predictedR3, test.data$FIP)
```

##FIP vs birthState

##Visualization
```{r}
ggplot(data=stateavgFIP, aes(x=birthState, y=FIP)) + geom_bar(stat="identity", color="blue", fill="white")
```

##models
##MLR
```{r}
set.seed(123)
training.samples <- stateFIP$FIP %>% createDataPartition(p = 0.8, list = FALSE)
train.data <- stateFIP[training.samples, ]
test.data <- stateFIP[-training.samples, ]
```

##SVR
```{r}
SVR_6_FIP <- svm(FIP ~  G + PO + A + E + DP + W + L + CG + SHO + SV + HR + BB + SO + BAOpp + WP + HBP + GF + ERA + birthState, data = train.data)

##Train Data
predicted.classes <- predict(SVR_6_FIP, train.data)
RMSE(predicted.classes, train.data$FIP)
R2(predicted.classes, train.data$FIP)

##Test Data
predicted.classes <- predict(SVR_6_FIP, test.data)

RMSE(predicted.classes, test.data$FIP)
R2(predicted.classes, test.data$FIP)
```

##DT
```{r}
DT_6_FIP <- rpart(FIP ~  G + PO + A + E + DP + W + L + CG + SHO + SV + HR + BB + SO + BAOpp + WP + HBP + GF + ERA + birthState, data = train.data, method = "anova")

plot(DT_6_FIP, uniform = TRUE,
     main = "Runs Decision Tree using Regression")
dev.off()
print(DT_6_FIP)
summary(DT_6_FIP)

##train
predictedR3 <- predict(DT_6_FIP, train.data, method = "anova")
RMSE(predictedR3, train.data$FIP)
R2(predictedR3, train.data$FIP)

##test
predictedR3 <- predict(DT_6_FIP, test.data, method = "anova")
RMSE(predictedR3, test.data$FIP)
R2(predictedR3, test.data$FIP)
```







## Question 4 What are the flaws of traditional baseball statistics?

## Visualization
```{r}
ggplot(data = Batting3, mapping = aes(x = yearID, y = wOBA)) + geom_point(color = "blue") + geom_smooth()
ggplot(data = Batting3, mapping = aes(x = yearID, y = BA)) + geom_point(color = "red") + geom_smooth()
```

## Modeling
```{r}
#Linear regression

Batting3[is.na(Batting3)] = 0
Batting3$BA[is.infinite(Batting$BA)] = 0
Batting3$wOBA[is.infinite(Batting$wOBA)] = 0

##model <-lm(wOBA ~ BA, data = Batting3)
##summary(model)

##view(Batting3)
#Multiple Linear Regression
##modelMult <- lm(wOBA ~ BA + G + AB + R + H + XBH + OBP + `2B` + `3B` + RBI + SB + CS + BB + SO + IBB+ HBP + SH, data = Batting3)
##vif(modelMult)
##summary(modelMult)
```

 

## Question 5 - IS SALARY A GOOD PREDICTOR OF AWARDS WON?

## Transformation

I removed the tie and notes variables from the dataframes, then I removed the teamID column since I didn't need it

```{r}
# Remove tie and notes variables from AwardsPlayers dataframe
Awards2 <- AwardsPlayers[ -c(5:6)]
kable(Awards2[1:15, ], caption = "Awards2 df after removal and filter")
Salaries2 <- Salaries[ -c(2)]
kable(Salaries2[1:15, ], caption = "Salaries2 df after removal and filter")
```

From here, I joined the two dataframes using a left join by their key and omitted any na values.

```{r}
AwardedSalaries <- left_join(Salaries2, Awards2, by = c("playerID","yearID"))
AwardedSalaries <- na.omit(AwardedSalaries)
kable(AwardedSalaries[1:15, ], caption = "Dataframe created from Left Join")
```

## Visualization 

To find any correlation among this data, I created a scatterplot visualization of the AwardedSalaries dataframe.

```{r}
ggplot(AwardedSalaries,aes(awardID, salary)) + 
  geom_beeswarm(color = "blue") + coord_flip()
```

For more exploration on this question, I created a boxplot as well.

```{r}
ggplot(AwardedSalaries,aes(awardID, salary)) +  geom_boxplot() + coord_flip()
```

## Testing for Correlation b/w Salaries and Awards

Furthermore, I conducted a test to compare the variance of Salaries for each Award to the overall variance of the AwardedSalaries dataframe. I broke the main dataframe into smaller ones and computed the variance for each, then composed them in a list.

```{r}
# Filter every award into its own variable
ALCSMVP <- AwardedSalaries %>% filter(awardID == "ALCS MVP")
AllStarMVP <- AwardedSalaries %>% filter(awardID == "All-Star Game MVP")
BabeRuthAward <- AwardedSalaries %>% filter(awardID == "Babe Ruth Award")
ComebackPlayer <- AwardedSalaries %>% filter(awardID == "Comeback Player of the Year")
CyYoungAward <- AwardedSalaries %>% filter(awardID == "Cy Young Award")
GoldGloveAward <- AwardedSalaries %>% filter(awardID == "Gold Glove")
HankAaronAward <- AwardedSalaries %>% filter(awardID == "Hank Aaron Award")
MVPAward <- AwardedSalaries %>% filter(awardID == "Most Valuable Player")
NLCSMVP <- AwardedSalaries %>% filter(awardID == "NLCS MVP")
PTC <- AwardedSalaries %>% filter(awardID == "Pitching Triple Crown")
RobertoAward <- AwardedSalaries %>% filter(awardID == "Roberto Clemente Award")
RolaidsRelief <- AwardedSalaries %>% filter(awardID == "Rolaids Relief Man Award")
RookieOfYear <- AwardedSalaries %>% filter(awardID == "Rookie of the Year")
SilverSlugger <- AwardedSalaries %>% filter(awardID == "Silver Slugger")
TripleCrown <- AwardedSalaries %>% filter(awardID == "Triple Crown")
TSN_AllStar <- AwardedSalaries %>% filter(awardID == "TSN All-Star")
TSN_Fireman <- AwardedSalaries %>% filter(awardID == "TSN Fireman of the Year")
TSN_Pitcher <- AwardedSalaries %>% filter(awardID == "TSN Pitcher of the Year")
TSN_Reliever <- AwardedSalaries %>% filter(awardID == "TSN Reliever of the Year")

# Determine variance in salaries of each award
ALCSMVPv = ALCSMVP$salary
ALCSMVP_Variance = var(ALCSMVPv)

ALLStarMVPv = AllStarMVP$salary
AllStarMVP_Variance = var(ALLStarMVPv)

BabeRuthAwardv = BabeRuthAward$salary
BabeRuthAward_Variance = var(BabeRuthAwardv)

ComebackPlayerv = ComebackPlayer$salary
ComebackPlayer_Variance = var(ComebackPlayerv)

CyYoungAwardv = CyYoungAward$salary
CyYoungAward_Variance = var(CyYoungAwardv)

GoldGloveAwardv = GoldGloveAward$salary
GoldGloveAward_Variance = var(GoldGloveAwardv)

HankAaronAwardv = HankAaronAward$salary
HankAaronAward_Variance = var(HankAaronAwardv)

MVPAwardv = MVPAward$salary
MVPAward_Variance = var(MVPAwardv)

NLCSMVPv = NLCSMVP$salary
NLCSMVP_Variance = var(NLCSMVPv)

PTCv = PTC$salary
PTC_Variance = var(PTCv)

RobertoAwardv = RobertoAward$salary
RobertoAward_Variance = var(RobertoAwardv)

RolaidsReliefv = RolaidsRelief$salary
RolaidsRelief_Variance = var(RolaidsReliefv)

RookieOfYearv = RookieOfYear$salary
RookieOfYear_Variance = var(RookieOfYearv)

SilverSluggerv = SilverSlugger$salary
SilverSlugger_Variance = var(SilverSluggerv)

TripleCrownv = TripleCrown$salary
TripleCrown_Variance = var(TripleCrownv)

TSN_AllStarv = TSN_AllStar$salary
TSN_AllStar_Variance = var(TSN_AllStarv)

TSN_Firemanv = TSN_Fireman$salary
TSN_Fireman_Variance = var(TSN_Firemanv)

TSN_Pitcherv = TSN_Pitcher$salary
TSN_Pitcher_Variance = var(TSN_Pitcherv)

TSN_Relieverv = TSN_Reliever$salary
TSN_Reliever_Variance = var(TSN_Relieverv)

# Create a table compiling salary variances by Award
Awards <- c("ALCSMVP","AllStarMVP","BabeRuthAward","ComebackPlayer",
            "CyYoungAward","GoldGloveAward","HankAaronAward","MVPAward",
            "NLCSMVP","PTC","RobertoAward","RolaidsRelief","RookieOfTheYear",
            "Silver Slugger","Triple Crown","TSN All-Star","TSN Fireman",
            "TSN Pitcher","TSN Reliever")

SalaryVariance <- c(ALCSMVP_Variance, AllStarMVP_Variance, BabeRuthAward_Variance, 
                    ComebackPlayer_Variance, CyYoungAward_Variance, 
                    GoldGloveAward_Variance, HankAaronAward_Variance, MVPAward_Variance, 
                    NLCSMVP_Variance, PTC_Variance, RobertoAward_Variance, RolaidsRelief_Variance, 
                    RookieOfYear_Variance, SilverSlugger_Variance, TripleCrown_Variance, 
                    TSN_AllStar_Variance, TSN_Fireman_Variance, TSN_Pitcher_Variance, 
                    TSN_Reliever_Variance)

AwSalVar <- data.frame(Awards, SalaryVariance, stringsAsFactors = FALSE)

kable(AwSalVar, caption = "Salary variance list by award")
```

Finally, I computed the overall variance of the AwardedSalaries dataframe and compared it to the individual variances of each award to reach the conclusion that Salary is not a good predictor of awards won.

```{r}
# Compute the variance of AwardedSalaries
AwardedSalariesv = AwardedSalaries$salary
AwardedSalaries_Variance = var(AwardedSalariesv)

kable(AwardedSalaries_Variance, caption = "Variance of AwardedSalaries dataframe")
```

## Question 6 Predicting Earned Runs (ER)
## Visualization
```{r}
ggplot(data = pitcher_stats_complete, mapping = aes(x = IP, y = ER)) + geom_point(color = "blue") + geom_smooth(color = "red")
ggplot(data = pitcher_stats_complete, mapping = aes(x = BB, y = ER)) + geom_point(color = "blue") + geom_smooth(color = "red")
```
## partitioning the data
```{r, echo=FALSE}
ER.ts <- pitcher_stats_complete$ER %>% createDataPartition(p = .8, list = FALSE)
ER.train <- pitcher_stats_complete[ER.ts,]
ER.test <- pitcher_stats_complete[-ER.ts,]
```
## Training

## MLR
```{r}
modelMLRER <- lm(ER ~ PO + A + CG + HR + BB + BAOpp + WP + HBP + SO,data = ER.train)
vif(modelMLRER)
summary(modelMLRER)
predictER <- predict(modelMLRER, ER.train)
RMSE(predictER, ER.train$ER)
R2(predictER, ER.train$ER)
```
## SVR
```{r}
modelSVRER <- svm(ER ~ PO + A + DP + L + CG + HR + BB + BAOpp + WP + HBP,data = ER.train)
predictedER <- predict(modelSVRER, ER.train)
RMSE(predictedER, ER.train$ER)
R2(predictedER, ER.train$ER)
```
## DT
```{r}
fitER <- rpart( ER ~ PO + A + DP + L + CG + HR + BB + BAOpp + WP + HBP, data = ER.train, method = "anova")
plot(fitER, uniform = TRUE,
     main = "Runs Decision Tree using Regression")
dev.off()
print(fitER)
predictedER2 <- predict(fitER, ER.train, method = "anova")
RMSE(predictedER, ER.train$ER)
R2(predictedER, ER.train$ER)
```
## Testing

## MLR
```{r}
predictER4 <- predict(modelMLRER, ER.test)
RMSE(predictER4, ER.test$ER)
R2(predictER4, ER.test$ER)
```
## SVR
```{r}
predictedER5 <- predict(modelSVRER,ER.test)
RMSE(predictedER5, ER.test$ER)
R2(predictedER5, ER.test$ER)
```
## DT
```{r}
predictedER6 <- predict(fitER, ER.test, method = "anova")
RMSE(predictedER6, ER.test$ER)
R2(predictedER6, ER.test$ER)
```

## Question 7 Regular Season OPS Vs Post Season OPS
```{r, echo=FALSE}
RvPOPS <-left_join(BattingPost3, Batting3,by = c("playerID", "yearID"))
RvPOPS <- RvPOPS %>% filter( G.y > 30)
RvPOPS <- RvPOPS %>% rename(Regular_Season_OPS = OPS.y)
RvPOPS <- RvPOPS %>% rename(Post_Season_OPS = OPS.x)
RvPOPS <- RvPOPS %>% rename(Regular_Season_AB = AB.y)
RvPOPS <- RvPOPS %>% rename(Post_Season_AB = AB.x)
RvPOPS <- na.omit(RvPOPS)
```
## Visualization
```{r}
ggplot(data = RvPOPS, mapping = aes(y = Post_Season_OPS, x = Regular_Season_OPS)) + geom_point(color = "blue") +geom_smooth()
```
## correlation test
```{r}
cor(RvPOPS$Regular_Season_OPS, RvPOPS$Post_Season_OPS)
```
## MLR
```{r}
model_PVRP <- lm(Post_Season_OPS ~ Regular_Season_OPS, data = RvPOPS)
summary(model_PVRP)
```

## Investigating the low R-Squared Values
```{r}
ggplot(data = RvPOPS, mapping=aes(x=Regular_Season_AB, y=Post_Season_AB))+geom_point(color = "blue")
```
## Conclusion

  When it comes to whether Salaries is a good predictor of awards won, based on the visualization and variance test we found that it is not. This could be due to Baseball as a sport paying players exorbitant amounts of money but also due to awards being given out for accomplishments and performance and not how much one makes, therefore the data and tests illustrate a reasonable result.